---
title: "Midterm Project Proposal"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Group Memebers
```{r, echo = F}
tibble(
  Name = c('Jianghui Lin', 'Zixu Wang', 'Jack Yan'),
  UNI = c('jl5172', 'zw2541', 'xy2395')
) %>% 
knitr::kable()
```

## Project Title 
Categorizing income level with census data

## Data Source

http://archive.ics.uci.edu/ml/datasets/Adult

## Data Description

The dataset includes the catagorical individual's income level and several predictors assiociated with the response. The extraction was done by Barry Becker from the 1994 Census database. Prediction task is to determine whether a person makes over 50K a year.  

__Responses:__

* individual's income level: >50K, <=50K.

__Predictors:__

* age: continuous.

* workclass: Private, Self-emp-not-inc, Self-emp-inc, Federal-gov, Local-gov, State-gov, Without-pay, Never-worked.

* fnlwgt: continuous.

* education: Bachelors, Some-college, 11th, HS-grad, Prof-school, Assoc-acdm, Assoc-voc, 9th, 7th-8th, 12th, Masters, 1st-4th, 10th, Doctorate, 5th-6th, Preschool.

* education-num: continuous.

* marital-status: Married-civ-spouse, Divorced, Never-married, Separated, Widowed, Married-spouse-absent, Married-AF-spouse.

* occupation: Tech-support, Craft-repair, Other-service, Sales, Exec-managerial, Prof-specialty, Handlers-cleaners, Machine-op-inspct, Adm-clerical, Farming-fishing, Transport-moving, Priv-house-serv, Protective-serv, Armed-Forces.

* relationship: Wife, Own-child, Husband, Not-in-family, Other-relative, Unmarried.

* race: White, Asian-Pac-Islander, Amer-Indian-Eskimo, Other, Black.

* sex: Female, Male.

* capital-gain: continuous.

* capital-loss: continuous.

* hours-per-week: continuous.

* native-country: United-States, Cambodia, England, Puerto-Rico, Canada, Germany, Outlying-US(Guam-USVI-etc), India, Japan, Greece, South, China, Cuba, Iran, Honduras, Philippines, Italy, Poland, Jamaica, Vietnam, Mexico, Portugal, Ireland, France, Dominican-Republic, Laos, Ecuador, Taiwan, Haiti, Columbia, Hungary, Guatemala, Nicaragua, Scotland, Thailand, Yugoslavia, El-Salvador, Trinadad&Tobago, Peru, Hong, Holand-Netherlands.

## Planned analyses

